DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems

### 2012-11-08 Paper_May_2012/main.tex
Paper: Sections aufgewertet

\subsection{Solving of PDEs}
\subsubsection{Stationary problems}\label{subsubsec:stationary problems}
For the start we consider a prototypical problem given by the solution of 
a stationary PDE, which we assume to be given in weak form.
In abstract notation it reads; find some $u$ in an appropriate set
such that
\begin{align}\label{eq:prototype_weak}
a(u)(\phi) = (f,\phi) \quad \forall \phi \in V,
\end{align}
with some suitable space $V$ for the test functions and a 
right hand side function $f$.
To illustrate this further, we take the stationary 
incompressible Navier-Stokes equations in a domain $\Omega \subset \R^d$, 
i.e., $u = (v,p)$ and $\phi = (\phi_u,\phi_p)$ then
\begin{align}\label{eq:ns}
a(u)(\phi) = \nu(\nabla v, \nabla\phi_u) + (v \nabla v,\phi_u) + (p, \nabla \cdot \phi_u) + (\nabla \cdot v ,\phi_p)
\end{align}
for all $\phi \in V = H^1_0(\Omega;\R^d) \times L^2_0(\Omega)$.
Here $(\cdot,\cdot)$ denotes the $L^2$-inner-product.
The solution will be searched for in $V + (u_d,0)$ where $u_d$ is some 
continuation of the given Dirichlet boundary data.
Then typically we do not only want to calculate $u$ but more likely we are 
interested in one (or several) functionals depending on the solution, i.e., some numbers $J(u)$.

@@

\paragraph{\texttt{main.cc}}
In this file we gather the basic components that are necessary for the
solution process of a (stationary) partial differential equation. The
structure of the program is basically the same for every equation under
consideration, what changes is typically just the \textit{kind} of  finite
element or the linear solver. Here, a key issue is that we only need to pick 
some pre-implemented routines from the \dope{} library rather than 
implementing new features.

@@

\paragraph{\texttt{localpde.h}}
We shall now give an overview to the implementation of the specific form of the PDE.

@@

\paragraph{\texttt{localfunctional.h}}
The implementation of the boundary-functional $J$ follows the outline of the computation of the cell-residual in \texttt{localpde.h}. However, there are two main differences. First, we do not have to specify the method \texttt{CellEquation}, but \texttt{BoundaryValue} instead. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face wise (for the faces that lie on $\Gamma_{circ}$). The second difference is that as we only want to evaluate a functional and not a residual-vector, we just loop over the quadrature points and not over the degrees of freedom on the face.
\begin{lstlisting}
 double
 BoundaryValue(const FaceDataContainer<2>& fdc)
  {
    //same as before, get n_q_points etc.
    fdc.GetFaceValuesState("state", ufacevalues);
    fdc.GetFaceGradsState("state", ufacegrads);
\end{lstlisting}
Notice that we get here the values of the computed solution by giving the string "state".
\begin{lstlisting}
    for(unsigned int q_point = 0; q_point < n_q_points; q_point++)
      erg += -1. * ( _nu * ufacegrads[q_point][0]
                * state_fe_face_values.normal_vector(q_point)
                + //other terms)        
                * state_fe_face_values.JxW(q_point);
    return erg;
  }
\end{lstlisting}

\subsubsection{Nonstationary problems}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l(\phi)$.
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\[
\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt
+ (v(0) - v^0, \phi(0))
= \int_I l(\phi) \, dt 
\]
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \bigl\{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R} \bigr\}.
\]
As for stationary problems, the goal is to compute
some target functional $J(\cdot)$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename Triangulation, typename FE, typename times>
      void SpaceTimeHandler(Triangulation &, FE, times&);
\end{lstlisting}

>>>>

\subsection{Solving of Stationary PDEs}
\label{subsubsec:stationary problems}
For the start we consider a prototypical problem given by the solution of 
a stationary PDE, which we assume to be given in weak form.
In abstract notation it reads; find some $u$ in an appropriate set
such that
\begin{align}\label{eq:prototype_weak}
a(u)(\phi) = (f,\phi) \quad \forall \phi \in V,
\end{align}
with some suitable space $V$ for the test functions and a 
right hand side function $f$.
To illustrate this further, we take the stationary 
incompressible Navier-Stokes equations in a domain $\Omega \subset \R^d$, 
i.e., $u = (v,p)$ and $\phi = (\phi_u,\phi_p)$ then
\begin{align}\label{eq:ns}
a(u)(\phi) = \nu(\nabla v, \nabla\phi_u) + (v \nabla v,\phi_u) + (p, \nabla \cdot \phi_u) + (\nabla \cdot v ,\phi_p)
\end{align}
for all $\phi \in V = H^1_0(\Omega;\R^d) \times L^2_0(\Omega)$.
Here $(\cdot,\cdot)$ denotes the $L^2$-inner-product.
The solution will be searched for in $V + (u_d,0)$ where $u_d$ is some 
continuation of the given Dirichlet boundary data.
Then typically we do not only want to calculate $u$ but more likely we are 
interested in one (or several) functionals depending on the solution, i.e., some numbers $J(u)$.

@@

\subsubsection{\texttt{main.cc}}
In this file we gather the basic components that are necessary for the
solution process of a (stationary) partial differential equation. The
structure of the program is basically the same for every equation under
consideration, what changes is typically just the \textit{kind} of  finite
element or the linear solver. Here, a key issue is that we only need to pick 
some pre-implemented routines from the \dope{} library rather than 
implementing new features.

@@

\subsubsection{\texttt{localpde.h}}
We shall now give an overview to the implementation of the specific form of the PDE.

@@

\subsubsection{\texttt{localfunctional.h}}
The implementation of the boundary-functional $J$ follows the outline of the computation of the cell-residual in \texttt{localpde.h}. However, there are two main differences. First, we do not have to specify the method \texttt{CellEquation}, but \texttt{BoundaryValue} instead. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face wise (for the faces that lie on $\Gamma_{circ}$). The second difference is that as we only want to evaluate a functional and not a residual-vector, we just loop over the quadrature points and not over the degrees of freedom on the face.
\begin{lstlisting}
 double
 BoundaryValue(const FaceDataContainer<2>& fdc)
  {
    //same as before, get n_q_points etc.
    fdc.GetFaceValuesState("state", ufacevalues);
    fdc.GetFaceGradsState("state", ufacegrads);
\end{lstlisting}
Notice that we get here the values of the computed solution by giving the string "state".
\begin{lstlisting}
    for(unsigned int q_point = 0; q_point < n_q_points; q_point++)
      erg += -1. * ( _nu * ufacegrads[q_point][0]
                * state_fe_face_values.normal_vector(q_point)
                + //other terms)        
                * state_fe_face_values.JxW(q_point);
    return erg;
  }
\end{lstlisting}

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l(\phi)$.
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\[
\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt
+ (v(0) - v^0, \phi(0))
= \int_I l(\phi) \, dt 
\]
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \bigl\{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R} \bigr\}.
\]
As for stationary problems, the goal is to compute
some target functional $J(\cdot)$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename Triangulation, typename FE, typename times>
      void SpaceTimeHandler(Triangulation &, FE, times&);
\end{lstlisting}
